Dynamic capabilities and competitive advantage

### 2019-10-21 article4_tommylars.rnw
after tommy meeting first draft

%% Extant literature on dynamic capabilities is battling with a fundamental conception and
%% prediction of the theory, namely if, when and how dynamic capabilities can lead to competitive
%% advantage.

Seminal contributions to the field of dynamic capabilities have argued for both direct
\cite{Kor2005,Eriksson2014,Schilke2014a} and indirect (ref) effects of dynamic
capabilities on competitive advantage, although increasingly the bulk of the literature
focuses on the latter \citep{Schilke2018}. This indirect link is typically understood as
dynamic capabilities modifying, extending or creating firm resources (helfat) (or
lower-level operational capabilities (Winter, collis)) and thus generating competetive
advantage (Eriksson and others). Furthermore, empirical evidence keep suggesting that both direct (ref)
and indirect (ref) links exist. This warrant the question of \emph{when} direct and
indirect links between dynamic capabilities and competitive advantage is most pronounced.

Moreover, firm resources or capabiltiies themselves are far from a homogenous set and
varies among other things with respect to their uniqueness, inimatbility and value
(barney). Important contributions have thus argued for a deeper insighi into these
intermediaries and their mechanism calling for a "deeper insight into the variety of mechanisms that
underlie the performance effects of capabilities.” (Schilke, 2014a: 199). 
as well as \emph{what} different intermediaries exis .

%% Extant literature on dynamic capabilities is battling with a fundamental conception and
%% prediction of the theory, namely if, when and how dynamic capabilities can lead to competitive
%% advantage. 
all items

       Path   LOW     t  HIGH     t
1 DC -> CA1 0.303 3.517 0.087 0.710
2   DC -> P 0.445 6.070 0.435 3.972
3  P -> CA1 0.273 3.174 0.308 3.164
> R
       Path   LOW     t  HIGH     t
1 DC -> CA1 0.438 6.161 0.134 0.988
2   DC -> R 0.400 5.661 0.549 8.033
3  R -> CA1 0.015 0.186 0.195 1.555
>

fewer items

      Path   LOW     t   HIGH      t
1 DC -> CA1 0.173 1.771 -0.017 -0.172
2   DC -> P 0.437 5.183  0.432  4.875
3  P -> CA1 0.329 3.646  0.353  3.731
> R
       Path   LOW     t  HIGH     t
1 DC -> CA1 0.292 2.891 0.049 0.438
2   DC -> R 0.142 1.287 0.370 4.267
3  R -> CA1 0.109 1.471 0.244 2.344

more items

       Path   LOW     t  HIGH     t
1 DC -> CA1 0.314 4.107 0.139 1.236
2   DC -> P 0.386 5.481 0.391 3.797
3  P -> CA1 0.291 3.596 0.292 3.011
> R
       Path   LOW     t  HIGH     t
1 DC -> CA1 0.439 6.095 0.163 1.122
2   DC -> R 0.428 5.547 0.579 7.972
3  R -> CA1 0.019 0.208 0.177 1.466

Seminal contributions to the field of dynamic capabilities have argued for both direct
\cite{Kor2005,Eriksson2014,Schilke2014a} and indirect (ref) effects of dynamic
capabilities on competitive advantage, although increasingly the bulk of the literature
focuses on the latter \citep{Schilke2018}. This indirect link is typically understood as
dynamic capabilities modifying, extending or creating firm resources (helfat) (or
lower-level operational capabilities (Winter, collis)) and thus generating competetive
advantage (Eriksson and others). Furthermore, empirical evidence keep suggesting that both direct (ref)
and indirect (ref) links exist. This warrant the question of \emph{when} and under which conditions direct and
indirect links between dynamic capabilities and competitive advantage is most pronounced.

Moreover, firm resources or capabiltiies themselves are far from a homogenous set and
varies among other things with respect to their uniqueness, inimatbility and value
(barney). Important contributions have thus argued for a deeper insighi into these
intermediaries and their mechanism calling for a "deeper insight into the variety of mechanisms that
underlie the performance effects of capabilities.” (Schilke, 2014a: 199). In other words,
adding to the question of when direct and indirect links are most pronounced, we would
also need to understand \emph{what} different forms of intermediatons work under these
aforementioned conidtions.

Still, important empirical contributions argue for a direct link although the emerginig
consensus of the literature argues for an indirect link. This warrants the research question:
"Under what conditions are dynamic capabilities working directly and indirectly on firm
competetive advangate".

We argue that the reason for these differing findings stems from a two distinctly
different factor; the /emph{environmental dynamism} facing the firm, and the /emph{nature
  of the mediating resource being employed}.

TWO observetions>
- Direct effect mediated away (disappears) in high velocity environments - Could be due to
conception of DC as simple rules and best practices. Do they change nature? Or is it that
they go from latent to constitutent and thus their impact becomes clearer

- The mean of mediation matter in low environments.

What happens? VRIO?
Routines are never VRIO
Processes are VRIO
VRIO concept*idea is based on normal levels of change. In dyn changes happens faster and
the modes of routines are more important because you need more rapid change and routinesa
re simpler models. By being changed by DC they can generate CVRIO simply because of their
rapid deployment. Processes on the other hand are equally important and do not get less
VRIO facing environmental dynamism. Why? Cause they are path dependent and unique to the
firm, however should expect them to be less important if speed of change is of
importance.

.

 and a certain consensus leans toward defining
the concept through the idea of a mediating effect through ordinary capabilities (winter,
helfat

“Moreover, we need deeper insight into the variety of mechanisms that
underlie the performance effects of capabilities.” (Schilke, 2014a: 199)

This,
we argue, is due to the way DC behave differently under high-velocity industries compared
to normal times, and goes back to the original contention between the seminal papers on EM
and TPS and their different understanding of the concept.

We argue that DC comes about working differently under different circumstances. As we cant
exactly measure and observe DC (teece) we can only observe its effects when called
upon. They are more likely to be called upon in high velocity environments when change is
most needed. Consecnuently Eisenhardts notion of constitutent DC becomes more pronounced
and visible under high velocity environments where DC turns to action working as
intended. In such conditions DC will indeed work to change underlying routines and
capabilities to enhance competetitieve advantage. However, what we usually observe in more
stable industries nd contexts, is DC as latent and hence more correlated with competetive
advantage cause it is related to the potential. Actual change of routines and capabilites
are less needed so the mediation is less pronounced and we will observe a more direct
effect cause other factors are playing into the CA explanation such as more
generalmanagement, resources and so on. As many of these factors normally working in
stable environments are left out in empirical models, the general idea of DC as a proxy
for a good firm and management becomes more prevalant and we see a direct effect as a
consequence. Proxy for a good firm is added and shows that being a good firm is more
important in slow enviromnetn cause much of the ability to govern ana manage abusiness is
captured in this variable in this conducion. DC is also more prevelant cause it captures a
similar side as a latent potential to generate CA and act more as aproxy for a good
firm. However, in high markets the impact of DC increases as it starts to work as a
distinct force of change changing underslying routines and capabilites. Here we also see
that the importance of other factors of a good firm diminishes because the usual metrics
are not important - path dependency matters less and the ability to change resources
matters increasingly and thus only ghrouhg mediation is CA achieved.

However, a difference in the type of underlying resource matter. And this also goes back
to the difference between simple rules of action and more complex routines. VRIO is
achieved through either direct effects from DC who is VRIO because of the path dependency
more pronounced in normal environments. 
evidence has pointed to a dirc

Noone has looked at how dc operates differently through different OC and how these
interact with the dynamism in the environment. Conventional wisdom holds that routines are
not able to ca[ture VRIO whereas processes can.

Extant literature on DC struggles with a fundamental conception and prediction of the
theory, namely if and when DC can lead to competitive advantage. This friction has been
dubbed the "elephant in the room" (peteraf) and been tracked back to the seminal papers of
Teece and Eisenhardt, as well as their their different theoretical antecedents of
evolutionary economics (ref) and behavioral theory (ref) respectively (ref Arndt
x2). Recently, distefaneo et al propopsed a theoretical integration of these varying
conceptions of DC theory. They suggest that latent and constitutent elements of DC
represent different constructs but are integrated in that they work differently in shaping
strategic change.

Constitutent DC based on simpler rules will not create CA directly. 
DC is often critizesd for being tautological and less focus has been placed on instances
when dynamism can lead to poorer competitive outcomes. 
Structural inertia, commitment \cite{schreyogg2007}. Under what conditions are changing
capabilities counterproducttive

The relationship between dynamic capabilities and competitive advantage has been one of
great contention in the extant literature \citep{Peteraf2013,Ambrosini2009}. From this
debate a richer understanding of how different dynamic capabilities (or functional forms
of them) work differently to change different types of firm resources
\citep{Helfat2015a,Teece2007,Ambrosini2009a} with implications for competitive advantage
\citep{Peteraf2013} has emerged. Increasingly the relationship between (1) dynamic
capabilities, (2) firm resources and routines, and competitive advantage is seen as a
complex system with interactions between different types of each 1 and 2. Many conceptual
contributions have excellently discussed varieties or functional forms of either dynamic
capabilities \citep{Ambrosini2009a,Helfat2015ad,DiStefano2014} or underlying resources or
routines \citep{Winter2003,Eisenhardt2000}. However, very few, if any, papers attempt to
integrate varieties at multiple levels simultaneously. By asking if dynamic capabilities
can lead to competitive advantage and how this occurs, one is far better off asking "when"
competitive advangate is an outcome of dynamic capabiliteis and its workings.

The definition of dynamic capabilities, albeit contestst, seems to be converging on
varieties of "the ability of firms to create, modicy and extend its resource base" (ref)
\citep{Schilke2018}. Meanwhile both dynamic capabilities as the "governing entity" and
resources as "the governed entity" takes different forms in the literature (ref). Dynamic
capabilities themsleves can roughly be divided into sets of simple rules governing
undelying resources, or more complex routines developed for the same purpose
\citep{Peteraf2013}. In other words, dynamic capabilities themselves can and cannot satisfy
the VRIO condition. This contention is throughly covered in the literature (refs) and
recent contribiutions have argued for interplays between simple and complex types of
dynamic capabilities foring a more complex and initmatble dynamic system
\citep{DiStefano2014}. Moreover, granted different forms of dynamic capabilities exist
their impact on the "governed entity", the underlying firm resources, should also be
varying.

To explore this we emply two radically differnet forms of dynamic capabilities / one
geared towards transforming exitsgint resources in order to exploit them to gain
competitive advantage (refs). The other form is geraed towards seizing new opportunities in order
to explore new business opportunitieis (refs). We argue that competitive advantage can
stem from each of these, but that the mechanism varies. Dynamic capabilities geared
towards transforming for exploitation ($DC_P$) is by and of itself the result of a more
complex set of routines configured in a way that is hard to imitate because of its path
dependent and evolutionary nature. This legacy from evolutinoary economics is stressed as
an important factor for creating VRIO resources (Ref peteraf). The stream of contributions
maintaining dynamic capabilities working through the object of "changing existing
capabilities" \citep{Teece1997,Zollo2002a,Winter2003,Zahra2006,Kale2002} also maintain
that

We argue that latent and constituted dynamic capabilites behave differently and work
differnetly trhough differents modes (action or wahtever the term in DiStefano er).

{\bf LAtent = complex process routines Teece
Constitutent = simpler, ad hoc - Eisenhardt}

We explore two different modes (natures) of dynamic capabilities and how they work
differently (action) to create competitive advantage. Or is it the object of action as
either resources or opportunities?

By introducing different measures of different types of DC (latent and constiuted) as well
as two different types of action (change existing or create new) we explore the workings
of the drive train model proposed by DiStefano suggesting a fuller and more dynamic view
on how latent and consittued DC works differently.

The working of latent:
Departing from the definintion of dynamic capabilities found in Teece (ref) and subsequent
contributions in this tradition

The working of cinstuted

We identify two ditintctly different noptions of dynamic capabilities both following well
known definitions byut weith different nature (ref distefano). We show that only the
latent dc can directly and indirectly create CA,whereas the constituent element nature of
DC only works through changing lower level routines.

Different parts of the drive train works on different types of underlying routines

Looks at competing hypotheses between Eisenhardt and teexe with respect to routinization. Finds that routininzation happens at DC level less than operational level?
We argue that this depends on the purpose *action of the Dc and the composition of the
focal operational capability.

Reviewing the literature reveals that researchers have tended to identify dynamic capabilities post hoc, inferring their existence from successful organizational out- comes such as profitability and growth, as prior definitions would suggest (Table

On the other hand we have dynamic capabilites geared toward sensing and seizing new
opportunities as part of an effort of exploration. Such dynamic capabilities are described
in the literature as themselves a set of more simple routines orchestrated in a manner
such that x`

Routinization at certain level, but depends on them being outward or inward looking.

How can DC lead to competitive advangate. departing from the original definition this
happens through the modification or aquisition of firm resources. But the resopurces
themselves can and cannot be a source of competitive advantage

At a minimum, in order for something to qualify as a capability, it must work in a reliable manner’ (Helfat and Peteraf, 2003: 999). Capabilities

The bulk of contributions focus
only on heterogeneity at the dynamic capabilities level, or at

W

work dDC offers a theory of how firms achieve
competitive advantage over time. Yet the DC theory suffers from contradictory claims about
competitive advantage as its outcome. On the one hand a stream of contributions argue that
DC will lead to competitive advantage because it operates by changing operational
capabilities in face of change and hence create strategic change (ref helfat). In this
view these changes result in complex and path dependent processes that will be
idiosyncratic to the firm and hard to imitate. Consequently, they can lead to competitive
advantage.

A contrary stream of literature argues that changes of operational capabilities over time
takes the form of routines and converges towards best practice (ref EM). Such best
practices are simpler and more observable routines that are both imitable and easy to
copy. Such firm resources are then not idiosyncratic and can consequently not be the
source of competitive advantage.

The way DC work differnelty throug changing two sets of operational capabilities = complex
and path dependent processes based capabilities, and simple rules/based routines.

These contradictory positions have been thoroughly discussed in earlier work (see
\cite{Peteraf2013} for a review). On the one hand DC takes the form of simple rules
shaping underlying routines. On the other hand they take the form of complex processes.

Regarding the action and the object ofthe action, simple rules are more likely to be
involved in developing something new in response to new opportunities, while complex
rou- tines are more concerned with changing existing resources and capabilities.

In a recent "state of the art" literature review on dynamic capabilities
\cite{Schilke2018} state that much of the attention and interest surrounding the theory is
that it offer a route to competitive advantage which they argue is "the virtual Holy Grail
of strategic management" (p. 393). And indeed a considerable body of the extant literature
concerns itself with exactly this path to competitive advantage (refs from
shcilke). Still, a discourse regarding the contingencies of the dynamic
capabilities-competitive advantage (DC-CA) nexus (?) has panned out in the literature,
most notably coined "the elephant in the room" \citep{Peteraf2013}: Under which conditions
will dynamic capabilities lead to competitive advantage?

\cite{Peteraf2013} argue that the seminal papers in the dynamic capabilities theory
tradition (\cite{Teece1997} and \cite{Eisenhardt2000}) differ considerably in their view
of to what extent the DC-CA nexus exist and under what conditions it is functional. A key
tenant of this discourse is related to the level of routinization and the ability to
satisfy the VRIO condition \citep{Barney1991a} of the resource-based view
\citep{Peteraf2013}.

In this paper we aim to contribute to dissecting the elephant by asking how does the DC-CA
nexus differ between firms and what determines these differences?

We do this by exploring two paths from dynamic capabilities to competitive advantage: the
\emph{evolutionary path} and the \emph{market path}. The former is thoroughly described
and explored in the literature (refs) and stems from the core idea of the dynamic
capabilities theorizing. In this path dynamic capabilities work to change the operating
routines of the firm and hence contributing to it's competitive advantage. We argue that
this path is more suited for less dynamic environments because of its path dependent
functioning.

The \emph{market path} works not through shaping operational capabilities and routines,
but by enabling the firm to act in factor markets for new capabilities and
resources. Hence, it is less path dependent and can be more rapidly
deployed. Consequently, it is more prevalent in dynamic environments.

This paper contributes to the extant literature on the link between dynamic capabilities
and competitive advantage in three ways. First, provides an empirical test to the
drive-train model of \cite{DiStefano2014} on how dynamic capabilities can operate both
quickly and with high frequency more directly on the strategic direction of the firm by
acting in factor markets for important resources, and also more slowly through the
evolutionary path of changing operational capabilities. Second, by conceptually clarifying
the different paths to competitive advantage by linking it closely to the role of
capability evolution and the VRIO condition. Finally, we.........

CLARIFIES THE ROLE OF THE ENVIRONMENT AND MEDIATIONS HAPPENING

%% How does dynamic capabilities lead to competitive advantage?
%% Dynamic capabilities are proposed to confer a competitive advantage by adding unique value to the firm through systematic change, which enhances operational efficiency and enables increased alignment with the environment (Di Stefano et al., 2014; Peteraf et al., 2013)

%% Not all organizations possess them (Collis, 1996); their path dependencies, intangibility, complexity, and organizational specificity make them hard to imitate (Gibson & Birkinshaw, 2004; Helfat & Winter, 2011); and few other means allow organizations to purposefully change on a continuous basis (Day & Wensley, 1988; Helfat et al., 2007).

%% Consistent with the theoretical positions of Eisenhardt and Martin (2000), Helfat and Peteraf (2003), Zahra et al. (2006), and Zott (2003), among others, these researchers have argued that dynamic capabilities’ immediate purpose is to change the resource base, and that this change in the resource base, in turn, explains performance variations. According to this argument, resource changes serve as mediators through which dynamic capabilities affect performance. We

%% One of the major unresolved issues in dynamic capabilities theory is to what extent and
%% how competitive advantage is the ultimate outcome of dynamic capabilities. This very idea
%% has

%% This paper asks two questions: to what extent does a direct lnk between DC and CA exist,
%% and what determines variations in the role of mediation between them?

%% The idea that Dc impacts CA is uncontroversial (not). Little is known about how this
%% happens. The theory holds that this happens through xreating extending and modifying the
%% resource base. YEt, the change in firm capabilities can themselves lead to CA due to their
%% evlutionary nature.

%% Variations in the share of mediation stems from variation sin the environment. When envdyn
%% is high firms need to act faster and hence tends to choose strategies including resource
%% orchestration buying in factor markets cause it tkes more time to develop OR.

%% In less dynamism the role of OR increases and leads to a significant proportion of CA
%% stemming throhg the mediating effect of OR

%% There is ongoing interest in (dynamic) capabilities and how they influence firm
%% performance \citep{Fainshmidt2016a,Teece2014}. Indeed, \emph{dynamic capabilities} (DC),
%% and its front runner, the resource-based view \citep{Barney2001,Helfat2007}, is now
%% considered one of the major theories in strategic management
%% \citep{Schilke2018}. Arguably, the key reason is that this theory clearly articulates how
%% firms can alter their operating routines through dynamic capabilities. Thus, a key
%% ingredient in this theory is that DCs can alter firms operating routines, with subsequent
%% performance implications.

We demonstrate that two paths indeed do exist and that their relative importance varies
with the degree of environmental dynamism facing the firm.

@@

%% We can use the drive train to determine hwo DC works differently under different levels of
%% dynamism. It is not that DC does not work under stable conditions as EM claims. Is only
%% that the effect runs stringer through the evolutionary path. Whereas the market path is
%% stringer under higher dynamism
This friction has been
dubbed the "elephant in the room" (peteraf) and been tracked back to the seminal papers of
Teece and Eisenhardt, as well as their their different theoretical antecedents of
evolutionary economics (ref) and behavioral theory (ref) respectively (ref Arndt
x2).

One contentious issue surrounding the DC theory is the extent to which differing views on
the nature of dynamic capabilities \citep{Peteraf2013} as well as their outcome. One
strand of contributions sees DC as routines that are simple in nature and more as best
practices to consider (EM). Thus, dynamic capabilities are more like behavioral patterns
and less idiosyncratic to the firm and will consequently not be able to satisfy the VRIO
condition for competitive advantage to emerge. Dynamic capabilities as such routines are
short lived, unstable and “substitutable” (EM: 1110) thus violating a key VRIN condition,
and they are more “more homogeneous (..) than is usually assumed” (EM: 1116). However, the
simplicity of simple rules also makes them faster do deploy and put in motion, albeit
potentially unstable. To the extent dynamic capabilities manifest themselves as simple
rules they cannot alone create competitive advantage, but could very well generate
strategic change by facilitating acquisition of resources in factor markets. One example
is the role of post-acquisition routines where routines to identify and deal with
acquisition partners are a source of generating strategic change (Eisenhardt and Sull
2001, ref). Another example is (see refs from Schilke 2018). The functioning of this
expression of dynamic capabilities are more latent and hence mainly observable ex
post. Hence we can see this as direct route to competitive advantage as this expression of
dynamic capabilities work directly in the factor market in capturing resources
needed for strategic change. We coin this direct connection as the \emph{market path}.

On the other hand, a stream of contributions see dynamic capabilities as evolutionary and
more complex routines (ref) or processes (ref) set to change underlying capabilities in
the face of changing conditions (TPS). This notion stems from evolutionary economics
\citep{Winter2003,Nelson1982} and the behavioral theory of the firm \citep{Cyert1963}
holding that path dependency and complex processes for changing underlying operating
routines are a source of competitive advantage for the firm. Along this indirect path
change is gradual and running through changes in underlying routines. We coin this the
\emph{evolutionary path}.

These paths from dynamic capabilities to competitive advantage are different expressions
of dynamic capabilities with different mediums of intermediation. The \emph{market path}
works in the market and his hence not typically altering lower order capabilities and
resources in the firm directly, but rather acquiring new ones. The source of competitive
advantage along this path stems from the interaction with the other, \emph{evolutionary
  path} in a dynamic bundle \citep{Peteraf2013}. Such interactions create a bundle of
capabilities and routines that are socially complex and hard to imitate
\citep[p. 320]{DiStefano2014}. Moreover, the interactions between the two paths of dynamic
capabilities is separable with respect to what intermediation exist. Consequently,
separate paths exist albeit a part of a dynamic bundle.

\cite{DiStefano2014} propose an integrated framework where simple routines and more
complex processes could contribute simultanously to the firms outcome, specifically its
competitive advantage. In this framework, coined the "organizational drivetrain" the
authors envision the movement of an organization as set in motion by a drivetrain system
not unlike that of a bicycle.

Building on the notion of an organizational drivetrain we argue that dynamic capabilities
can indeed impact the organizational movement or change through two distinct but
interconnected mechanisms (i.e. the crank set and the rear shifter). The two mechanisms
consitutting the drivetrain is consistent with our notion of \emph{evolutionary path} and the \emph{market path}. The
evolutionary path is instigated by more complex processes setting in motiong a chance in
the lower order (operational) routines \cite{Collis1994,Winter2003}. In the drivetrain
metaphor this path corresponds to change stemming from the front crank set of thich there
are fewer gears, but with larger impact. This path is slower
and yields a more idiosyncratic set of organizational routines which is very much in line
with the idea put forth in the seminal founding papers \citep{Teece1997,Winter2003}.

Seeing the bike rider and her efforts to handle the levers of control (i.e. the gear
shifter) as the dynamic capability, and consider it not only as a management effort, but
that of the organization as a whole. The process of changing front gears are more profound
(i.e. evolutionary) and costly, and it is for larger and slower strategic
changes. Operating with simple routines in factor markets, however, is quicker and the
rider does this on a more frequent basis with less effort.

H1: Dynamic capabilities works to create competitive advantage through two distinctly
different paths.

\subsection{The two paths of dynamic capabilities}
The objective of strategic change is to enhance a firm’s competitive advantage
(ref). Doing so over time makes for the evolutionary fitness of the firm
\citep{Helfat2007}. When firms are developing capabilities for strategic change through
creating, extending and modifying the resource base this means making decisions to develop
operational routines. Such routines can in turn stem from an inherent learning process
evolutionary distinct to the firm. Operating routines can thus lead to competitive
advantage because of their evolutionary character \citep{Nelson1982,Winter2003}. To the
extent operating routines stems from idiosyncratic evolutionary process of individual
firms rather than being the result of investments in factor markets, they can satisfy the
VRIO condition and thus be a source of competitive advantage. In such instances, dynamic
capabilities work through operating routines in creating competitive advantage. This
indirect impact on competitive advantage is very much at the core of the DC literature
(ref). Firms with high dynamic capabilities will be able to better change operating
routines to facilitate strategic change in response to changing conditions facing the
firm. Dynamic capabilities then creates VRIO operating routines and hence competitive
advantage. When this happens, firms changes their operating routines in a evolutionary
making operating routines themselves OR itself satisfies the VRIO condition.

\emph{H2: Dynamic capabilities has an indirect effect on competitive advantage through
  changing the operating routines of the firm.}

>>>>

%% Extant literature on dynamic capabilities is battling with a fundamental conception and
%% prediction of the theory, namely if, when and how dynamic capabilities can lead to competitive
%% advantage. 
%% This paper is about how apsobritve capacity can come about as strategic posture
%https://arrow.dit.ie/cgi/viewcontent.cgi?article=1009&context=buschmancon

% paper about the DC for MNC in india an dimpact on CA
%https://journals.sagepub.com/doi/abs/10.1177/0971890717701781

The presence of a direct path from dynamic capabilities and competetive advantage is
somewhat puzzling. It eiter suggests ommited importrant mediating factors such as
resources, or that dynamic capabilities can act outside "creating, extending and modifying
the firm resources base". A growing consensus on the definition as well as the
theoretical mechanisms underpinning dynamic capabilities suggest that dc work indirectly
towards strategic change by changing firm resources. Hence, the presence of direct effects
seems increasingly odd. Still, seminal contributions to the field of dynamic capabilities have argued for both direct
\cite{Kor2005,Eriksson2014,Schilke2014a} and indirect (ref) effects of dynamic
capabilities on competitive advantage, although increasingly the bulk of the literature
focuses on the latter \citep{Schilke2018}.

 Resource orchestration is defined as "blablbalbla" (helfat)

This indirect link is typically understood as dynamic capabilities modifying, extending or
creating firm resources (helfat) (or lower-level operational capabilities (Winter,
collis)) and thus generating competitive advantage (Eriksson and others).The direct path,
however, is theoretically less explicitly dealt with in the extant literature although
empirical evidence keep suggesting that both direct (ref) and indirect (ref) links exist.

In this paper we explore to what extent a direct link between dynamic capabilities and
competitive advantage work, and under which conditions it is most pronounced. Specifically
the research question is:

{\bf How is a direct link between dynamic capabilities and competitive advantage working
  and under which conditions does it come into play or not?}

We argue that direct link exist as a response to the multilevel notion of strategy itself
and captures strategic change not typically associated with the resource based
view. Furthermore, we show that the prevalence of the direct link is a function of the
nature of underlying resources dynamic capabilities are aimed at "creating, extending and
modifying", specifically if they are valuable and idiosyncratic to the firm or not, and
the environment facing the firm.

Our findings indicate that the direct path is less prevalent in very dynamic environments
where the indirect path becomes far more important. Furthermore, the effect of dynamism is
partly contingent on the nature of the firm resources suggesting a more complex
understanding of how imitable and simple resources can play a role in generating
competetive advantage.

Our contributions are twofold. First, we explore different mediators by explicitly
modeling a causal chain with different types of underlying resources, as well as
under different environments. This is explicitly called for in the literature as "only
four percent of the articles in our sample explicitly examined possible causal mechanisms,
suggesting a significant gap in the literature" (Schilke 2018). This work sheds light on
the contingencies and boundary conditions of the link between dynamic capabilities and the
"holy grail of strategy" (Schilke et al), competetive advantage.

Second, we explore a theoretical explanation for why a direct link between dynamic
capabiliites and competetive advantage may exist. By leaning on empirical evidence
suggesting a multitude of explanations for performance differentials between firms, we
suggest that dyanmic capabilities can act outside the theorized path of "resource
orchestration" through strateic efforts such as industry positioning (porter) or
institutional leverage (Peng).

%%%%%%%%%%%%%%5

%%%%%%%%%%%%%%

%% This warrant the question of \emph{when}
%% direct and indirect links between dynamic capabilities and competitive advantage is most
%% pronounced.

%% Moreover, firm resources or capabiltiies themselves are far from a homogenous set and
%% varies among other things with respect to their uniqueness, inimatbility and value
%% (barney). Important contributions have thus argued for a deeper insighi into these
%% intermediaries and their mechanism calling for a "deeper insight into the variety of mechanisms that
%% underlie the performance effects of capabilities.” (Schilke, 2014a: 199). 
%% as well as \emph{what} different intermediaries exist.

%% This paper investigates the boundary conditions of the VRIO condition and how dynamic
%% capabilities and resources interact to create firm competitive advantage. We ask the
%% following research question:
%% Under which conditions are dynamic capabilities able to yield competietive advantage
%% through resource orchestration and what are the different outcomes of different types of resources.

%% We find thatthe dynamism surrounding hte firm represent an important boundary condition as
%% suggested in parts of the literature. However, DC consinutes to generate competetive
%% advnatage also in high-velocity environments, but doing increasingly so indirectly through
%% resource orchestration. Furthermore, we find that non-vrio resources are more likely to
%% yield competetive power in high velocity environments, and that vrio-resources tend to
%% maintain their CA under any condition.

%% We demonstrate that two paths indeed do exist and that their relative importance varies
%% with the degree of environmental dynamism facing the firm.

%% Under what cionditions are the direct path and indirect path most prevalant? Under which condi

%% The direct

@@

%{\bf How is a direct link between dynamic capabiliites and competetive advantage working
%  and under which conditions does it come into play or not?}

In the face of a growing consensus on the definition (ref) and outcome (ref) of dynamic
capabilities as working through mediators, the presence of direct effects argued by others
(ref) seems puzzling. However, the broader empirical literature on the antecedents of firm
performance offers an interesting point of departure. This literature concerns itself with
the relative importance of factors at the firm-level on the one hand, and those at the
industry level at the other.

The debate over the relative role of different levels of factors in accounting for the
major share of the variance in firms performance has been of particular importance in the
field of strategic management, most notably in the Strategic Management Journal
(M. A. Fitza 2014; Hansen and Wernerfelt 1989; McGahan and Porter 1997; Rumelt
1991). Overall, the accumulated empirical evidence show that firm factors matter far more
for firm performance than industry factors (McGahan and Porter 2002). These stylized facts
have been used as empirical support for the importance of idiosyncratic firm resource and
capabilities; leading up too recent formulation of the resource and capability based
theorizing (e.g. (Barney 1991; Teece, Pisano, and Shuen 1997)). However, the dominance of
the firm-level explanations have been partly refuted by later scholars finding the sum of
indirect and direct effects from industry to be of more importance than previously
indicated by (Mcgahan etc). Moreover, the industry effect is also more time invariant
(i.e. more stable) compared to firm level effects. Recent findings also suggest that there
are considerable interaction effects between industry and the firm indicating that
industry membership works directly at performance as well as indirectly through affecting
the firm. However, little or no explanations are made for the mechanisms of this
interaction.In other words, industry factors matter alas less than firm-level factors, and
the indirect link from industry-level through firms are in need of further development.

One way of looking at this is to consider the dynamic capabilities of the firm. Although
an increasing consensus focuses on these dynamic capabilities as the ability to create,
extend and modify firm resources in the face of change, there still looms an important
stream of literature suggesting that dynamic capabilities can materialize as management
actions (ref eisenhardt, peteraf, helfat). A management action aimed at strategic change
will very often entail "resource orchestration" as broadly discussed in the literature
(Helfat etc). But, strategic change are also found to come about from adapting and
changing the positioning of the firm within its industry (ref) or simply refocusing its
focal industry (ref). In other words, when a firm with excellent dynamic capabiliites
senses and seizes opportunities within its domain, the implementation of strategic change
will often both entail a considerable share of "asset orchestration" (helfat etc) as well
as management actions directed at positioning within its industry (ref). One way to
illustrate this is to consider Microsofts strategic change from operating system to cloud
provider. CEO Satya Nadela and his team was able to restart the business by creating,
extending and modifying the resources of the firm, including new skills, routiens and
hardware, as a response to a growing demand for cloud computing (ref nadellas
book). Meanwhile, his vision and formulated strategy also direclty put the company into a
new industry, or at the very least reshaped the position they were already in. While
Microsofts dynamic capabiliites worked significantly through "resource orchestration",
the sensing and seizing of the management also ordered a direct re-positioning at the same
time.

Moreover, The link between firm resources and competitive advantage is determined by the
extent to which resources are valuable and difficult to imitate \citep{Barney1991a}. This
principle is also used to evaluate how higher order capabilities (e.g. dynamic
capabilities \citep{Winter2003,Collis1994,Teece2007}. Hence, for a direct link to exist
the dynamic capabilities of the firm will have to be valuable and difficult to
imitate. This is an area of certain contention in the literature \citep{Peteraf2013} which
we get back to below, but an increasing majority is arguing that dynamic capabilities
reflect more complex processes that co-evolve with responses to market changes over time
and hence creating idiosyncratic capabilities that are unique, valuable and hard to copy
\citep{Helfat2007,Arndt2018}. Consequently, they are likely to lead to competitive
advantage.

In most cases such direct effects between dynamic capabilities and strategic change
resulting in competitive advantage, are combined with the indirect effects (i.e. resource
orchestration). To test this proposition empirically is hard, but one hypothesis following
from this is related to the role of mediators. if a direct path does not exist, the
relationship between dynamic capabilities and competitive advantage would be fully
mediated by the proper set of firm resources.

\emph{H1: The existence of a direct path from dynamic capabilities to competitive advantage
suggest that full mediation is not present}

Even if the dynamic capabilities of the firm themselves are not valuable and inimitable as
needed to create competitive advantage, the underlying resources of the firm may very well
be. Such resources in the form of operating routines or capabilities can in turn stem from
an inherent learning process evolutionary distinct to the firm. Operating routines can
thus lead to competitive advantage because of their evolutionary character
\citep{Nelson1982,Winter2003}. To the extent operating routines stems from idiosyncratic
evolutionary process of individual firms rather than being the result of best practices,
they can satisfy the VRIO condition and thus be a source of competitive advantage.  Hence,
through "resource orchestration" a firm with non-valuable dynamic capabilities can achieve
competitive advantage through the indirect path. The flip side of this is that a valuable
dynamic capability would not be able to create competitive advantage by orchestrating
non-valuable resources. Consequently, the mediation effects of underlying resources is
dependent on the nature of the resource - it being valuable and inimitable or not.

\emph{H2a: When dynamic capabilities are being mediated by VRIO resources the firm can
  achieve competitive advantage}

\emph{H2a: When dynamic capabilities are being mediated by VRIO resources the firm can not
  achieve competitive advantage}

DEspite a growing consensus on the nature of dynamic capabilites, a serious contention is
presen tin the literature \citep{Peteraf2013}. This stems from different notions of
dynamic capabilities in the two "founding" papers of the theory, \cite{Eisenhardt2000} and
\cite{Teece1997}.

One strand of contributions sees DC as routines that are simple in nature and more as best
practices to consider \citep{Eisenhardt2000}. Thus, dynamic capabilities are more like
behavioral patterns and less idiosyncratic to the firm and will consequently not be able
to satisfy the VRIO condition for competitive advantage to emerge. Dynamic capabilities as
such routines are short lived, unstable and “substitutable” (ibid, p 1110) thus violating
a key VRIO condition, and they are more “more homogeneous (..) than is usually assumed”
(ibid, 1116). However, the simplicity of simple rules also makes them faster do deploy and
put in motion, albeit potentially unstable. To the extent dynamic capabilities manifest
themselves as simple rules they cannot alone create competitive advantage, but could very
well generate strategic change by facilitating the process of resource orchestration.. In
other words, even if dynamic capabilities play out as simple rules that are not in
themselves satisfying the VRIO condition, the way they work through creating, extending
and modifying firm resources, can still yield competitive advantage through orchestrating
VRIO resources.

On the other hand, \cite{Teece1997} and others \cite{Helfat2007} argue that dynamic
capabilities will rather invariably be VRIO due to the evolutionary nature of their
development \citep{Arndt2018,Peteraf2013}. Arguing that dynamic capabilities reflect more
complex processes that co-evolve with responses to market changes over time and hence
creating idiosyncratic capabilities that are unique, valuable and hard to copy
\citep{Helfat2007,Arndt2018}. Consequently, they are likely to satisfy the VRIO condition
and thus leading to competitive advantage.
